The Private School Network: Recruiting Visits to Private High Schools by Public and Private Universities

### 2023-11-09 r_r/manuscript_rihe.Rmd
working on revisions

---
title: "The Private School Network: Recruiting Visits to Private High Schools by Public and Private Universities"
#author: 
#  - Ozan Jaquette
#  - Crystal Han
#  - Irma Castañeda
bibliography: ./../other/burd_recruiting_chapter.bib
output: 
  bookdown::pdf_document2:
    toc: FALSE
    pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
csl: ./../other/apa.csl
urlcolor: blue
header-includes:
      - \usepackage{pdflscape}
      - \usepackage{geometry}
      - \usepackage{floatrow}
      - \usepackage{setspace}
      #- \doublespacing
      - \floatsetup{capposition=top}
fontsize: 12pt
---

@@

ABSTRACT

Few empirical analyses of privatization compare public and private universities on a behavior associated with private universities. This manuscript analyzes off-campus recruiting visits to private high schools, which were made in 2017 by a convenience sample of 15 public research universities and 14 selective, private universities. Sociological scholarship on the relationship between high schools and universities conceives of off-campus recruiting visits as an indicator of a social relation. Therefore, we utilize social network concepts and methods to compare the recruiting networks of public and private universities. With respect to scale (research question 1), most public research universities in our sample made a disproportionate number of visits to private (out-of-state) high schools, though not as many as private universities. With respect to characteristics of visited schools (RQ2), both public and private universities tended to visit predominantly white schools in their home geographic region and the South. Public universities tended to visit lower ranked private high schools than private universities. Surprisingly, several public research universities visited sectarian private high schools at a rate similar to sectarian private universities. With respect to overlap (RQ3), descriptive statistics and community detection methods revealed substantial overlap in the recruiting networks of public and private universities.

@@

In the early twenty-first century, "privatization" received substantial attention from scholars of higher education [e.g., @RN921; @RN2101; @RN1118; @RN1577; @RN2111]. @RN4543 argue that scholarship on privatization waned because of a lack of conceptual and empirical clarity. One strand of scholarship conceptualizes privatization as a process whereby declines in government funding cause public institutions to grow alternative revenue sources. Public research universities responded, for example, by pursuing research commercialization and nonresident tuition revenue, and donative resources [e.g., @RN3753; @RN2101; @RN1644]. A second strand within the privatization literature conceptualizes privatization as a process whereby public institutions become more similar -- in terms of characteristics or behavior -- to for-profit or private non-profit organizations [@RN4543]. A weakness of this literature is that few empirical studies analyze public institutions on a behavior associated with private institutions. Moreover, few studies directly compare public and private institutions on the same behavioral dimension.

We argue that scholarship on recruiting behavior can shed light on the extent to which public universities are engaging in behaviors associated with private universities. Admissions counselors from colleges and universities visit local high schools in order to meet prospective students and maintain relationships with guidance counselors at feeder schools [@RN3519]. @RN4758 analyzed off-campus recruiting visits made in 2017 by 15 public research universities, finding that 12 of these universities made more visits to *out-of-state* high schools than *in-state* high schools. These out-of-state visits focused on public schools in affluent, predominantly white communities. Whereas @RN4758 primarily analyzed visits to public high schools, this chapter focuses on visits to private high schools. More specifically, we investigate how visits to private high schools by public research universities compare to visits to private high schools by selective private universities.

Our focus on recruiting visits to private high schools is motivated by seminal scholarship within the sociology of education. Sociologist Max Weber is credited for the insight that education has a "dual character," on one hand, serving *social mobility* by redistributing access to privileged positions and, on the other, serving *social reproduction* by granting privileged families disproportionate access to the educational institutions that confer pathways to privileged positions [@RN2377; @RN1133].

The mission of selective private colleges and universities is more squarely associated with social reproduction rather than social mobility. Historically, these institutions depend on patronage from philanthropic interests -- in connection with business and religious interests -- and on wealthy households who provide tuition revenue and donations [@RN4517]. In turn, selective private institutions contribute to social reproduction by providing access to their patrons. The "Chetty" data on parental income by college show that the overwhelming majority of students at selective private institutions have parental income in the top quintile [@RN4722; @RN4723]. A significant caveat is that religiously affiliated universities may prioritize enrollment opportunities for students of the same faith regardless of wealth, but even here the mission is to benefit families from a particular cultural status group [@RN4517].

The traditional, albeit mythologized, public good mission of public research universities is tied to social mobility and development of the state [@RN2269; @RN921]. Public research universities are designated the unique responsibility of educating the future civic, professional, and business leaders of the state. Quoting 19th century University of Michigan President James Angell, these institutions contribute to social mobility by providing "an uncommon education for the common man" [as cited in @RN3608, p. 279] who could not afford tuition at elite private institutions.

We argue that the recruiting behavior of colleges and universities is an indicator of organizational enrollment priorities and, in turn, the extent to which enrollment priorities serve social mobility or social reproduction. Recruiting visits to private high schools are consistent with the social reproduction mission of selective private universities, regardless of whether the university is targeting a particular high school because of its academic reputation, religious affiliation, or the affluence of its student body. By contrast, for public research universities, visits to private high schools -- particularly out-of-state schools -- are antithetical to the mission of social mobility for high-achieving, low-income state residents.

This manuscript compares public research universities to selective private universities on the dimension of recruiting visits to private high schools, a behavior associated with the social reproduction orientation of selective private universities. Our primary goal is to learn about public research universities, using the behavior of private universities as a benchmark to develop more thoughtful insights about public research universities. We analyze recruiting visits made in the 2017 calendar year by a sample of 15 public research universities and 14 selective private universities. 
<!-- and 12 selective private liberal arts colleges. -->

<!--
Note: The data for this book chapter will be slightly different due to using 2017-18 CCD data plus modified criteria (no longer excluding HS w/ certain school types) - but % is still similar

UC Berkeley: 171 (42%) of 404 out-of-state HS visits are to private schools
CU Boulder: 381 (38%) of 1008 out-of-state HS visits are to private schools
-->

Little is known about recruiting at private high schools by public universities. Therefore, we conduct descriptive analyses that seek to identify basic patterns and relationships. Because a recruiting visit to the same private high school is an indicator of a relationship between two actors, we also utilize social network methods in order to compare the recruiting networks of public universities to those of private universities. Analyses are informed by the following three research questions:

1. How does the scale of visits to private high schools by public research universities compare to the scale of visits to private high schools by selective private universities?
1. How do the characteristics of private high schools visited by public research universities compare to the characteristics of private high schools visited by selective private universities?
1. To what extent do public research universities and selective private universities visit overlapping sets of private high schools?

The remainder of the manuscript is organized as follows. First, we situate off-campus recruiting visits within the broader process of recruiting students and we review scholarship on the relationship between high schools and universities, focusing on scholarship from sociology. This literature motivates off-campus recruiting visits as an indicator of a social relation. Second, we introduce concepts from social network analysis and motivate empirical research questions. Third, we describe data and methods. Fourth, we present results. Finally, we discuss implications for scholarship.

<!-- SHORT PARAGRAPH GIVING OUTLINE OF CHAPTER -->

<!--  ORIGINAL INTRODUCTION

Prior to the pandemic, admissions counselors from colleges and universities traveled to local high schools each year in order to meet prospective students and maintain relationships with guidance counselors at feeder schools [@RN3519]. @RN4758 analyzed off-campus recruiting visits made in 2017 by 15 public research universities. Surprisingly, 12 of the 15 universities made more visits to *out-of-state* high schools than *in-state* high schools. These out-of-state visits focused on public schools in affluent, predominantly white communities. Additionally, most public research universities in our sample made a disproportionate number of visits to out-of-state private high schools. For the University of California-Berkeley, for example, 155 (43%) of the 360 out-of-state high school visits were to private high schools. For the University of Colorado-Boulder, 321 (36%) of the 901 out-of-state high school visits were to private high schools.

Whereas @RN4758 primarily analyzed visits to public high schools, this chapter focuses on visits to private high schools. More specifically, we investigate how visits to private high schools by public research universities compare to visits to private high schools by selective private universities.

## Social mobility, social reproduction, and privatization

The sociologist Max Weber is credited for the insight that education has a "dual character," on one hand, serving *social mobility* by redistributing access to privileged positions and, on the other, serving *social reproduction* by granting privileged families disproportionate access to the educational institutions that confer pathways to privileged positions [@RN2377].

The mission of selective private colleges and universities is more squarely associated with social reproduction rather than social mobility. Historically, these institutions depend on patronage from philanthropic interests -- in connection with business and religious interests -- and on wealthy households who provide tuition revenue and donations [@RN4517]. In turn, selective private institutions contribute to social reproduction by providing access to their patrons. The "Chetty" data on parental income by college show that the overwhelming majority of students at selective private institutions have parental income in the top quintile [@RN4722; @RN4723]. A significant caveat is that religiously affiliated universities may prioritize enrollment opportunities for students of the same faith regardless of wealth, but even here the mission is to benefit families from a particular cultural status group.

The traditional, albeit mythologized, public good mission of public research universities is tied to social mobility and development of the state [@RN2269; @RN921]. Public research universities are designated the unique responsibility of educating the future civic, professional, and business leaders of the state. Quoting 19th century University of Michigan President James Angell, these institutions contribute to social mobility by providing "an uncommon education for the common man" [as cited in @RN3608, p. 279] who could not afford tuition at elite private institutions.

We argue that the recruiting behavior of colleges and universities is an indicator of organizational enrollment priorities and, in turn, the extent to which enrollment priorities serve social mobility or social reproduction. Recruiting visits to private high schools are consistent with the social reproduction mission of selective private universities, regardless of whether the university is targeting a particular high school because of its academic reputation, religious affiliation, or the affluence of its student body. By contrast, for public research universities, visits to private high schools -- particularly out-of-state schools -- are antithetical to the mission of social mobility for high-achieving, low-income state residents.

Analyses of recruiting visits to private high schools can contribute novel insights about the "privatization" of public research universities. Within the scholarly literature on privatization, one strand of scholarship conceptualizes privatization as a process whereby declines in government funding cause public institutions to grow alternative revenue sources, particularly revenue from tuition and research commercialization [@RN4543; @RN2101]. Of relevance to this chapter, public research universities responded to declines in state appropriations by growing enrollment from nonresident students [@RN3753], who typically pay two- to three-times more than resident students. A second strand within the privatization literature conceptualizes privatization as a process whereby public institutions become more similar -- in terms of characteristics or behavior -- to for-profit or private non-profit organizations [@RN4543]. A weakness of this literature is that few empirical studies analyze public institutions on a behavior associated with private institutions. [FURTHER, FEW STUDIES COMPARE THE BEHAVIOR OF PUBLIC UNIVERSITIES TO THAT OF PRIVATE UNIVERSITIES ON THAT BEHAVIOR.]

Therefore, this chapter compares public research universities to selective private universities on the dimension of recruiting visits to private high schools, a behavior associated with the social reproduction orientation of selective private universities. Our primary goal is to learn about public research universities, using the behavior of private universities as a benchmark to develop more thoughtful insights about public research universities.  We analyze recruiting visits made in the 2017 calendar year by a sample of 15 public research universities and 14 selective private universities.

Because little is known about recruiting at private high schools by public universities, we conduct descriptive analyses that seek to identify basic patterns and relationships. Because a recruiting visit to the same private high school is an indicator of a relationship between two actors, we also utilize social network methods in order to compare the recruiting networks of public universities to those of private universities. Analyses are informed by the following three research questions:

1. How does the scale of visits to private high schools by public research universities compare to the scale of visits to private high schools by selective private universities?
1. How do the characteristics of private high schools visited by public research universities compare to the characteristics of private high schools visited by selective private universities?
1. To what extent do public research universities and selective private universities visit overlapping sets of private high schools?

-->

# What do we know about off-campus recruiting visits

This section provides context for our analyses.  First, we situate off-campus recruiting within the broader set of marketing and recruiting interventions in higher education and review what market research says about off-campus recruiting. Second, we review empirical scholarship from sociology that considers off-campus recruiting, emphasizing scholarship that considers the relational aspects of recruiting visits and scholarship that considers private high schools.

## Situating off-campus recruiting within enrollment management

The "enrollment funnel" -- depicted in Figure \@ref(fig:em-funnel) -- is a conceptual heuristic that identifies stages in the student recruitment process (prospects, inquiries, applicants, accepted applicants, and enrolled students). "Prospects" are "all the potential students you would want to attract to your institution" [@RN4322]. "Inquiries" are prospects that contact the institution, including those who respond to a solicitation and those who reach out on their own. The enrollment management industry uses the enrollment funnel to inform marketing and recruiting interventions that target specific stages. Most scholarship on enrollment management focuses on the final stages of the enrollment funnel, specifically which applicants are admitted and the use of financial aid "leveraging" to convert admits to enrollees [e.g., @RN3536; @RN3544; @RN3523; @RN4131; @RN2241; @RN3564; @RN1948]. By contrast, the enrollment management industry expends substantial resources on marketing/recruiting activities that target earlier stages of the enrollment funnel [@RN4664]. <!--[SHOULD WE TRY TO ADD SOME CONCRETE NUMBERS ON $ SPENDING AT PARTICULAR UNIVERSITIES?]-->

INSERT FIGURE \@ref(fig:em-funnel) ABOUT HERE

Institutions identify undergraduate "prospects" by purchasing "student lists" -- containing contact, demographic, and academic achievement information -- from College Board, ACT, and other vendors. Universities control which prospects are contained within a list by filtering on criteria such as zip code, test scores, and high school GPA. Once prospects and inquiries are identified, they are targeted with remote and face-to-face recruiting interventions designed to solicit applications and deepen engagement. Remote recruiting interventions include postcards, brochures, emails, text messages, and targeted social media. Face-to-face recruiting interventions include on-campus visits by prospects and off-campus recruiting visits by admissions representatives to high schools, community colleges, college fairs, etc. Institutions utilize advertising (traditional and digital) and social media to solicit inquiries and to create positive "buzz" amongst prospects [@RN4411]<!--[CITE OTHER]-->. Given the rise in "stealth applicants" who do not inquire before applying (Dupaul & Harris, 2012), advertising and social media enables universities to tell their story to prospects who do not wish to be contacted.

What do we know about off-campus recruiting from previous research? As is true for most aspects of enrollment management, much of what we know about off-campus recruiting comes from consulting firms (e.g., Ruffalo-Noel Levitz, EAB), professional associations (e.g., NACAC), and from practitioner-oriented publications. Market research describes off-campus recruiting visits as a means of identifying prospects and deepening engagement with prospects already being targeted through mail/email [@RN4323; @RN4402; @RN4664].

@RN4402 documents the self-reported efficacy of marketing/recruiting interventions. For the median private 4-yr institution, off-campus visits were the second highest source of inquiries (after student list purchases), accounting for 17% of inquiries. Off-campus visits were tied with student list purchases as the highest source of enrollees, accounting for 18% of enrollees. For the median public institution, off-campus visits accounted for 19% of inquiries (second only to student list purchases) and accounted for 16% of enrollees (ranked third after stealth applicants and on-campus visits).

With respect to expenditure, Table \@ref(tab:ug-recruit-budget), reproduced from @RN4664, shows the percentage of undergraduate recruitment budget allocated to different marketing and recruiting activities.^[Table drawn directly from @RN4664 Figure 9, which is based on a convenience sample of 45 four-year non-profit institutions and 21 four-year public institutions.] The average public university spent 16% of its recruiting budget on off-campus recruiting visits, compared to 12% of its budget on purchasing student lists and 17% of its budget on "prospective student communications" (e.g., mail, email, text, social media) targeting prospects and inquiries. An emergent trend over the past decade -- partially a response to public universities seeking nonresident students -- has been the growth of "regional recruiters" who target specific metropolitan areas in the US and abroad [@RN4021]. These regional recruiters may be college/university employees or they may be independent contractors who live in the metropolitan area they recruit.

<!-- QUESTIONS ABOUT THE RNL 2020 COST OF RECRUITING AN UNDERGRADUATE STUDENT REPORT, FIGURE 9

What does "recruitment publications" include?
what does "prospective student communications" include?
what does "event" include? can I assume this is on-campus events only cuz "travel" is a separate category?
what does "web services and digital advertising" include? Does this include social media?
-->

## Empirical scholarship from sociology

A sociological case-study literature develops helpful insights about the mechanisms and functions of off-campus recruiting visits [CITE].

-->

Empirical academic scholarship on off-campus recruiting is mostly limited to a modest number of sociological case studies, which typically analyze off-campus recruiting as part of a broader analysis of enrollment management [e.g., @RN4538; @RN4324; @RN3519; @RN4520; @RN4407]. @RN4324 analyzes visits from the perspective of students at two racially and socioeconomically diverse public high schools. @RN4324 found that high school visits -- including college fairs, instant decision events, and small-group representative visits -- influenced where students applied and where they enrolled. This finding was strongest for first-generation students and underrepresented students of color, who often reported that "school counselors had low expectations for them and were too quick to suggest that they attend community college" (p. XXX). This trust vacuum created an opportunity for colleges because these students were drawn to colleges that connected with them and made them feel wanted. For example, @RN4324 describes a high-achieving, first-generation, African American student who was admitted by a highly selective liberal arts college but chose to attend a less selective college that "seemed to want him more" (p. XXX). By contrast, affluent students with college-educated parents were less taken by such overtures and more concerned with college prestige.

@RN3519, in an ethnography of the admissions office at a selective private liberal arts college, highlights the relational function of off-campus recruiting visits. Stevens [-@RN3519, p. 54] states that "the College's reputation and the quality of its applicant pool are dependent upon its connections with high schools nationwide." Therefore, during the autumn "travel season," admissions officers visit selected high schools across the country "to spread word of the institution and maintain relationships with guidance counselors" (p. 53-54). The College tended to visit the same "feeder" schools year after year because recruiting depends on long-term relationships with high schools. The high schools they visited tend to be affluent schools -- in particular, private schools -- that enroll high-achieving students who can afford tuition and have the resources and motivation to host a successful visit. Whereas @RN4402 highlights the effect of recruiting visits on inquiries and enrollees, findings from @RN3519 suggest that the College may have valued recruiting visits primarily as a means of maintaining relationships with guidance counselors. From this perspective, recruiting visits may affect outcomes such as inquiries, applications, and matriculation through their effect on high school guidance counselors. The logic is that a guidance counselor who views a college favorably will steer students to the college.

@RN3519 and Khan [-@RN4321;-@RN4407] are centrally important to our study because they analyze relationships between college admissions counselors and high school guidance counselors.

-->

>>>>

---
title: "The Private School Network: Recruiting Visits to Private High Schools by Public and Private Universities"
#author: 
#  - Ozan Jaquette
#  - Crystal Han
#  - Irma Castañeda
bibliography: ./../other/burd_recruiting_chapter.bib
output: 
  bookdown::pdf_document2:
    toc: FALSE
    pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
csl: ./../other/apa.csl
urlcolor: blue
header-includes:
      - \usepackage{pdflscape}
      - \usepackage{geometry}
      - \usepackage{floatrow}
      - \usepackage{setspace}
      - \usepackage{fancyhdr}
      - \usepackage{titlesec}
      #- \doublespacing
      - \floatsetup{capposition=top}
      - \newcommand{\beginsupplement}{
      - \setcounter{table}{0}  
      - \renewcommand{\thetable}{S\arabic{table}} 
      - \setcounter{figure}{0} 
      - \renewcommand{\thefigure}{S\arabic{figure}}}
fontsize: 12pt
---

@@

@@

*During the 2017 calendar year, we tracked off-campus recruiting visits made by 14 selective private universities, 12 selective private liberal arts colleges, and 15 public research universities.^[NAME THE UNIVERSITIES] We are still trying to figure out what happened.*
\bigskip

Autumn is college recruiting visit season at Harpeth Hall, a non-sectarian private school for girls located in the affluent Green Hills neighborhood of Nashville, Tennessee. Founded in 1951, Harpeth is an "A+" school ranked 337 in the nation according to the 2020 *Best Private High Schools in America* by Niche. Enrollment patterns at Nashville-area private schools reflect broader regional patterns, whereby “efforts by the courts to desegregate public schools in the South resulted in White flight to private schools” [@RN4504, p. 14]. In Fall 2017, about 85% of Harpeth students were white and less than 5% were Black or Latinx (Author calculations).

From September 6th through October 17th, Harpeth received recruiting visits from 11 private universitiesin our sample (Northwestern, Boston College, Tufts, Villanova, Case Western Reserve, University of Denvor, Emory, Tulane, Baylor, Texas Christian University), and three private liberal arts colleges (Harvey Mudd, Scrips, and Sewanee). Previous scholarship observes strong connections between private schools and selective private universities [@RN3519; @RN4407]. However, Harpeth also received visits from six out-of-state public research universities, the University of Cincinatti, University of South Carolina, the University of Alabama, University of Arkansas, University of Georgia, and CU Boulder.

The nonsectarian University School of Nashville was founded in 1975 and is located next door to Vanderbilt University. It is an "A+" school, ranked 127th in the nation, and is more racially diverse than Harpeth (64% white, 11.5% Black, 3.5% Latinx, 10.0% Asian). From September 11th 2017 through October 18th, the University School received visits from nearly the same set of 11 private universities that visited Harpeth (Notre Dame visited and Baylor did not) and five liberal arts colleges (including Williams and Middlebury), but only two public research universities from our sample (University of Cincinatti, CU Boulder).

Father Ryan High School is a Catholic school founded in 1925. Father Ryan is predominantly white (86%, 7% Black or Latinx) and received an "A" grade from Niche, meaning it is not ranked in the top 1,000 US private schools. In fall 2017, Father Ryan received visits from nine of the 14 private universities in our sample (including all the Catholic universities) and the same six public universities that visited Harpeth Hall, including two visits by the University of Alabama. Meanwhile, Christ Presbyterian Academy -- also not ranked in the top 1,000 private schools -- received visits from five private universities in our sample (including the three Christian universities Baylor, Texas Christian University, and Southern Methodist University). It also received visits from the University of Cincinatti, the University of South Carolina, the University of Alabama, University of Arkansas, and CU Boulder. Christ Presbyterian Academy was founded in 1985 and 94% of its students are white.

Theories of organizational behavior suggest that university recruiting behaviors -- those that require significant expenditure -- are indicators of organizational enrollment priorities. Large, public-facing organizations are beset by multiple, competing pressures from the external environment, but organizations have difficulty pursuing multiple goals [@RN531; @RN959]. Thus, organizations respond by symbolically adopting some goals and substantively adopting others [@RN513]. Under symbolic adoption, organizations adopt policies and structures that send signals to external constituents but do not substantively affect resource allocation inside the the "technical level" of the organization.^[Contingency theory states that organizations consist of "technical levels" responsible for processing inputs and producing outputs and "managerial levels" responsible for directing the technical level and tending to demands from the external environment [@RN531]. The technical level achieves effectiveness by focusing on a small, stable set of goals.] Under substantive adoption, organizations direct resources in the technical level towards achieving the goal. Off-campus recruiting visits to local high schools and college fairs was the top-ranked recruiting expenditure item for both public and private universities, according to @RN4664. Therefore, analyses of which schools a university visits -- and which schools they ignore -- can yield insights about university enrollment goals.

@RN4758 analyzed recruiting visits to public high schools made by 15 public research universities in 2017. the authors found that 12 of the 15 universities made more visits to out-of-state schools than in-state schools. These out-of-state visits focused on affluent, predominantly white public schools. Set against the scholarship about declining state appropriations and growing nonresident enrollment [@RN3753; @RN3685], these results suggest that many public research universities prioritize enrollment from wealthy out-of-state students who can afford to pay nonresident tuition price.

Within scholarship on social mobility and social stratification, a multidisciplinary literature analyzes the relationship between private school and college [CITE]. This research observes strong links -- and student flows -- between private high schools and selective private universities [CITE]. To date, scholarship on nonresident enrollment by public universities has progressed in isolation from scholarship on links between between private school and college. We argue that these two stories have become connected because public research universities increasingly target out-of-state private schools, schools that have been the domain of selective private universities.

<!-- 
This study compares recruiting visits to private high schools by public research universities and recruiting visits to private high schools by selective private universities. 
We argue that recruiting visits are an indicator of enrollment goals -- the university wants to enroll some students from the school, the school wants some students to enroll at the university. 
-->

This study analyzes recruiting visits to private high schools made in the 2017 calendar year by a sample of 15 public research universities and 14 selective private universities.  <!-- @RN3519 finds that travel to local high schools is vital for maintaining relationships with guidance counselors at "feeder schools."--> We conceptualize recruiting visits as an indication the university and the school have a social relationship. This conceptualization motivates the use of social network methods, which analyze networks defined by "network ties" (i.e., visits) between actors (universities and schools). Comparing the recruiting networks of public research universities and selective private universities yields novel insights about the competition for affluent, "full-pay" students. Analyses are informed by three research questions:

1. How does the scale of visits to private high schools by public research universities compare to the scale of visits to private high schools by selective private universities?
    - ALTERNATIVE: How does the scale of visits to private high schools vary for the public research universities and selective private universities in our sample?
1. To what extent do public research universities and selective private universities visit overlapping sets of private high schools?
1. How do the characteristics of private high schools visited by public research universities compare to the characteristics of private high schools visited by selective private universities?

Both private and public universities in our sample made a disproportionate number of visits to private high schools. For public research universities, visited private schools were much more likely to be out-of-state than visited public schools. Although universities visited a large number of private schools in their home region, most universities also visited a large number of private schools in the South. Private schools that received visits were predominantly white. Although partially a function of our analysis sample, public universities tended to visit lower ranked private high schools than private universities. Surprisingly, several public universities visited sectarian private high schools at a rate similar to sectarian private universities. Network analyses revealed substantial overlap in the recruiting networks of public and private research universities.

This manuscript is practically important because it provides insight about which schools universities view as "feeder schools" and -- to a limited extent -- the characteristics of students they want to enroll. Prior research finds that university recruiting efforts affect the enrollment decisions of students [CITE]. In turn, the composition of enrolled students affects the campus culture experienced by students.  Scholarship and the popular press reports that underrepresented minority students and low-income students at public flagship universities feel isolated amidst affluent, predominantly white student bodies. Racial and socioeconomic isolation negatively affects student development outcomes for isolated students and for the entire student body [CITE]. State policymakers and institutional leaders should recognize that the enrollment composition at public research universities is not the consequence of exogenous student demand, but rather the consequence of recruiting efforts that seek to grow nonresident enrollment by systematically targeting affluent, predominantly white public high schools [@RN4758] and affluent, predominantly white private high schools.
[NOTE: THESE IMPLICATIONS HAVE TO DO ABOUT TARGETING PRIVATE SCHOOLS BUT NOT TARGETING SAME SET OF SCHOOLS TARGETED BY SELECTIVE PRIVATE UNIVERSITIES]

[PUT THIS IN THE DISCUSSION SECTION] Finally When both private and public universities are targeting private schools, competition for these students is likely to increase. This could be an explanation for why many private universities are struggling to meet enrollment goals

student bodies that predonominantly affluent, white student bodies.

# Literature Review

## Nonresident Enrollment at Public Universities

We stiuate this study vis-a-vis scholarship on enrollment management behavior by public research universities and particularly the pursuit of affluent nonresident students, a population also targeted by selective private universities. The so-called "iron triangle" of enrollment management states that universities pursue the broad enrollment goals of access, academic profile, and revenue generation [@RN3524; @RN2772]. Because resources are scarce, the imagery of the iron triangle suggests that the pursuit of one enrollment goal may involve trade-offs with other goals. @RN2772[p. 221] state that “most enrollment management policies\ldots do not advance all three objectives; instead they lead to gains in some areas and declines in others.” Scholarship argues that selective private universities prioritize academic profile and revenue generation, with funds generated from tuition and donation feeding the arms race for prestige [@RN3522; @RN1549; @RN3519]. Historically, public research universities prioritize enrollment of high-performing state residents, often with an emphasis on social mobility for students who cannot afford private universities [@RN3608]. However, scholarship on privatization argues the public research universities increasingly emphasizing tuition revenue and rankings and de-emphasize access for state residents [@RN1577; @RN1118].

<!-- PREVIOUS PARAGRAPH
We situate this study vis-a-vis scholarship on nonresident enrollment at public research universities, which lies within the broader literature on "privatization" [CITE]. Against the backdrop of many states disinvesting in higher education, the privatization literature examines the ways that public universities start behaving like private universities by emphasizing revenue generation and the pursuit of prestige in research and enrollment management behaviors. The so-called "iron triangle" of enrollment management states that universities pursue the three broad enrollment goals of access, academic profile, and revenue generation [CITE]. "Because resources are scarce, the imagery of the iron triangle suggests that the pursuit of one enrollment goal may involve tradeoffs with other goals. Scholarship on privatization expressed concern that the public research universities were emphasizing tuition revenue and rankings and deemphasizing access for state residents.-->

@RN3753 draw from resource dependence theory to conceptualize the relationship between state appropriations and nonresident enrollment. Organizations are dependent on external resources that are important for organizational survival/stability and cannot be easily replaced. Public universities depend on state appropriations but these funds became increasingly uncertain in the 1980s and 1990s [@RN2271]. Resource dependence theory states that one response to resource decline or uncertainty is to seek alternative, substitute resources [CITE]. Tuition revenue is a potential substitute for state appropriations. Most states cap resident tuition price, but not nonresident tuition price. Therefore, @RN3753 argued that public universities have a financial incentive to grow nonresident enrollment in response to declines in state appropriations. Analyzing panel data from 2002-03 to 2012-13, @RN3753 found that a 10% decline in state appropriations was associated with a 2.7% increase in nonresident enrollment at all public universities and a 5.0% increase in nonresident enrollment at research-extensive universities.

Other studies of nonresident enrollment examine university enrollment composition. Using data from NPSAS 2011-12, @RN3685 show that compared to resident students at public research universities, nonresident students generate higher net tuition revenue, are more affluent, are less likely to be Pell recipients and are less likely to be Black or LatinX. Institution-level panel models reveal that growth in the share of nonresident students was associated with a decline in the share of Pell recipients and a decline in the share of Black and LatinX students. @RN4290 examine whether growth in the number of nonresident students causes resident enrollment to decline. They find that nonresident enrollment "crowds out" resident enrollment at prestigious public research universities, but not at less prestigious public research universities.

How did public research universities go about attracting all these nonresident students? One mechanism is financial aid, a lever to increase applications and affect matriculation decisions. Many public research universities have developed institutional aid policies for nonresident admits, typically offering larger awards for higher achieving students [@RN3762; @RN2287; @RN1791].<!--@RN2535 shows that nonresident students are less sensitive to sticker price than resident students. Because nonresident prospects may be considering several options -- in-state, out-of-state, public, private -- they are sensitive to the generosity of institutional aid offers [CITE]. For example, @RN3762 analyze the effect of a University of Iowa institutional merit aid program for nonresident admits that offered higher awards for admits with higher academic achievement. Regression discontinuity analyses found that the program increased the probability of enrollment, particularly for high achieving students. These results suggest that high-achieving nonresident students face many good opportunities and require substantial institutional aid to be convinced to enroll. -->

Earlier in the "enrollment funnel," universities utilize marketing and recruiting interventions to attract nonresident applicants. One widely utilized intervention is off-campus recruiting visits by admissions counselors to local high schools. @RN4758 analyzed off-campus recruiting visits to public high schools by 15 public research universities. Recruiting visits to in-state high schools tended exhibit modest racial and socioeconomic bias. However, 12 of the 15 universities made more visits to out-of-state high schools than in-state high schools. These out-of-state visits focused on public schools in affluent, predominantly white communities. @RN4934 conducted geo-spatial analyses of out-of-state recruiting visits to Los Angeles and Dallas by four public research universities. She found that universities engage in "recruitment redlining -- the circuitous avoidance of predominantly Black and Latinx communities along recruiting visit paths" (p. X).  Although @RN4758 and @RN4934 focused on visits to public high schools, prior research has not considered visits to private schools by public research universities. For example, what is the scale of these visits and what are the characteristics of visited schools?

## Linkages Between Private Schools and Private Universities

Separate from scholarship on nonresident enrollment at public universities, scholarship from economics and sociology analyzes linkages between private schools and colleges.

@RN4504<!-- use data from several national surveys to --> set the stage by describing long-term trends in private school enrollment. The percentage of American (elementary school) students attending private school declined from 15% in the mid-1960s to 10% in the mid-1970s, and declined gradually from 11% in 1999 to 9% in 2015. The private school enrollment rate for high-income families and for low-income families has remained relatively stable over time at about 17% and 4%, respectively. The enrollment rate for middle-income families declined from 12% in 1968 to 6% in 2013. This decline was driven by shrinking Catholic school enrollment, which represented 89% of private school enrollment in 1965 and 42% of enrollment in 2013. Moreover, contemporary Catholic school students tend to be more affluent compared to prior decades. There has been a long-term increase in private school enrollment in the South, with middle-income families increasingly attending Christian schools and high-income families increasingly attending nonsectarian schools. With respect to race, 11% of white students attended private school in 2013, compared to 5% of Black students, and 3% of Hispanic students.
<!-- Private school enrollment rates are higher for families located in urban areas than those in suburban areas.
 However, for high income families, rates of private school enrollment do not differ substantially across race. -->

A robust empirical literature<!-- from sociology and economics  --> finds that attending private school positively affects college access and college selectivity [e.g.,] [@RN4642; @RN4596; @RN4634; @RN4599]. Scholarship from the 1980s and 1990s focused on the effects of attending Catholic school [@RN4937; @RN786; @RN4936; @RN4938]. More recent scholarship tends to find that attending a nonsectarian private school has larger effects on college access than attending a Catholic school [@RN4935; @RN4614;@RN4625].

The High School Longitudinal Survey of 2009 (HSLS) tracks the postsecondary destinations of students who were in ninth grade in 2009. Simple descriptive statistics are eye-opening (Author calculations).^[Analyses consist students who met the following conditions: completed the 2012 follow-up survey; obtained high school transcript data; and completed the 2016 followup survey. The survey weight variable used was W4W1W2W3STU. After weighting, these 14,020 (unweighted) students represent a population of 3.93 million U.S. ninth graders in 2009. Data on college rankings come from the 2020 U.S. News \& World Report *Best Colleges*. Data on high school control come from the HSLS school file and from the Private School Survey.] About 93% of students attended a public high school and 7% attended a private high school, including 3.6% attending a Catholic school, 1.6% attending a nonsectarian school, and 1.8% attending a Christian school. Of public high school students, 5.0% attending a public university ranked in the top 100 by US News and World Report (USNWR) and 1.9% attended a private university ranked in the top 100. Of private high school students, 10.9% attended a top 100 public university and 16.8% attended a top 100 private university. For Catholic school students, 12.8% attended a top 100 public university and 11.7% attended a top 100 private university, while for Christian schools these percentages were 8.2% and 8.2%, respectively. However, for nonsectarian private school students, 11.3% attended a top 100 public university and 39.8 percent attended a top 100 private university. Focusing on the institutional composition of private universities in USNWR top 100, 39.8% attended a private high school (compared to 7% of all high school students who attended a private high school) and 21.2% attended a nonsectarian private high school (compared to 1.6% of all high school students who attended a nonsectarian private high school).

Scholarship has been particularly interested in the relationship between attending a prestigious private school and attending an ultra selective university [@RN4647; @RN4643; @RN4647; @RN4589]. @RN4935 acquired unprecedented data -- application files and admissions decisions linked to income tax records  -- to investigate why high income families are over-represented at "Ivy Plus" colleges. Conditioning on SAT and ACT scores, applications from high-income families (top 1%) received the same academic rating as applications from lower-income families. However, applications from high-income families received higher non-academic ratings (extracurricular activities, leadership traits) then applications from lower-income families. In turn, the higher non-academic ratings for high-income applicants was driven by the fact that these students tended to attend elite private high schools that devote substantial resources to teacher recommendations and guidance counselor letters.
